---
title: "index.Rmd"
author: "Sem Beentjes"
date: '2023-02-27'
output:
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: bootstrap
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(spotifyr)
library(tidyverse)
library(flexdashboard)
library(plotly)
library(Cairo)

thebends <- get_playlist_audio_features("", "3YGj8blWCJoTWfEngyKQAM?si=9752c01ddc294055")

okcomputer <- get_playlist_audio_features("", "4oskn4uMBsPa2iC9oElNpb?si=1a03352822d140ac")


bends_vs_computer <-
  bind_rows(
    thebends |> mutate(category = "The Bends"),
    okcomputer |> mutate(category = "OK Computer"))

bendscomputer_plot <- ggplot(bends_vs_computer, aes(valence, tempo, size = loudness, color = mode)) +
  geom_point() +
  facet_wrap(~ category) +
  theme_light() +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.50, 1),
    minor_breaks = NULL) +
  labs(
    x = "Valence",
    y = "Tempo",
    colour = "Mode",
    size = "Loudness"
  ) +
  scale_size_continuous(      
    guide = "none"           
  ) +
  scale_color_continuous(
    guide = "none")
  
interactive_plot_bendsvscomputer <- ggplotly(bendscomputer_plot)

kid_a <- get_playlist_audio_features("", "7h8g0HLDFVc4YOnZuyn83s?si=e6e0b0179cce46d5")
amnesiac <- get_playlist_audio_features("", "48VcKYl1HHCLyQI24jiwpn?si=26e3d2c26e6945b8")


kida_vs_amnesiac <-
  bind_rows(
    kid_a |> mutate(category = "Kid A"),
    amnesiac |> mutate(category = "Amnesiac"))

kida_amnesiac_histo <- ggplot(kida_vs_amnesiac, aes(valence, fill = key_name)) +
  geom_histogram() +
  facet_wrap(~ category) +
  theme_light()

ggplotly(kida_amnesiac_histo)

kida_amnesiac_point <- ggplot(kida_vs_amnesiac, aes(valence, acousticness, fill = key_name, color = key_name)) +
  geom_point() +
  facet_wrap(~ category) +
  theme_light()

ggplotly(kida_amnesiac_point)


```

### Kid A and Amnesiac, histogram

The next two albums i'd like to compare, are Kid A (2000) and Amnesiac (2001). These two albums are interesting, because they were recorded during the same period and were originally meant to be one big album together. But eventually, it was decided that it would be better to turn them into two separate albums, which lead to Kid A receiving a lot of praise worldwide, since it was (again) very experimental and different. Amnesiac was a but less popular when it was released, almost seen as a slight step back towards more traditional rock stuff (though there are still lots of experimental songs on it). So, let's compare these albums and see if we can conclude anything about the keys used in the albums.

```{r, echo=FALSE}

ggplotly(kida_amnesiac_histo)

```

***

In this histogram, where the color of the bars indicated the key any given song is in, and the count indicated the amount of songs with the same (or very similar) valence score. Noteworthy is that the amount of low-valence songs is higher in Amnesiac, which would support the overall idea that Amnesiac is less experimental than Kid A.

***

Also, I think it is interesting that Amnesiac's two high-valence songs are in the keys of G and C, while those keys are also used by two of the four high-valence songs in Kid A. And while G seems to be a commonly used key in both albums, C is only used twice, both in high-valence songs. So, just on the basis of these two albums, you might say Radiohead likes to experiment in the key of C.

### Kid A and Amnesiac, second graph

After the histogram, there is still a part of the comparison that we haven't really looked at, and that is whether it's possible to tell from Spotify analysis features if these albums would've made a good album if merged together.

```{r, echo=FALSE}

ggplotly(kida_amnesiac_point)

```

***

In this plot, I've compared Kid A and Amnesiac on basis of valence vs acousticness, with the color once again being the key. I wanted to use the color for key again, mostly to compare those two songs in the key of C with high-valence scores, and now I can see that they are not only close in valence, but also in acousticness. So, Radiohead likes to experiment with acoustic, high-valence music when in the key of C (on basis of these two albums).

***

The acousticness is very similar in both albums, with a very comparable spread on the y-axis of this plot. This supports the idea that these albums were originally meant to be released together, as combining the albums wouldn't really affect the acousticness balance of both albums. One thing that can be said is that Kid A's high-acousticness songs are slightly higher on the acousticness scale. 

### Chroma and Timbre analysis of "Paranoid Android"

Because Paranoid Android is one of the biggest hits of Rahiohead's OK Computer album, as well as being one of their most experimental songs, being over 6 minutes long and not really having a traditional or expected structure, I decided to look at this song and draw a spectogram for both the chroma and the timbre.
The spectogram for the chroma is very interesting, as it shows that the experimental features of the song heavily influence the key changes, and that there are more than four different keys that the song is at different points.
The timbre spectogram is less interesting, as no real conclusions can be drawn from it. The timbre of the song is fairly consistent and, besides the first guitar solo, there are no real deviations during the song.

### Introduction

Radiohead is an English alternative-rock band known mostly for their use of experimental soundscapes and abtract lyricism knowadays, but this wasn't always the case. Before the release of their third commercial album "OK Computer" in 1997, Radiohead made more guitar-centered, lyrically introspective music. OK Computer is largely believed to have laid the groundwork for their later, more experimental albums such as "Kid A" and "Hail to the Thief". But was this album really so groundbreaking for the band, or were there signs of more experimental interests within the bands earlier albums, when analysing them in Spotify? Or is it possible that Spotify altered some labels/genres because of the image Radiohead had already acquired before Spotify released in the first place? Looking at the audio feature for each of their albums, I hope to be able to showcase the change in Radioheads music throughout the years.

### Comparing the albums The Bends (1995) and OK Computer (1997), with Valence on the x-axis and Tempo on the y-axis


```{r, echo=FALSE}

interactive_plot_bendsvscomputer

```

***

In the provided graph, two Radiohead albums are compared to eachother, The Bends (1995) and OK Computer (1997). While The Bends is mostly seen as a rock album, without that much groundbreaking/experimental elements, and a heavy focus on guitars. But OK Computer was seen as groundbreaking at the time, and was a lot more focused on experimental soundscapes and abstract lyrics. 

***

In the graph, the albums are compared on tempo and validity, with the size of the dots representing loudness and the color representing mode. One interesting thing to note is that the validity ratings of OK Computer are visibly lower than the ones on The Bends. This means that, according to Spotify, OK Computer is a more sad album compared to The Bends. 


***

Also noteworthy is that the tempo on OK Computer seems to be lower than the tempo on The Bends, which supports the notion that The Bends was more of a rock album than OK Computer. Lastly, the amount of songs in a minor vs major key are not significally different between the albums, with The Bends having 3 out of their 12 songs in a minor key and OK Computer having 4 out of their 12 songs being in minor key.